A Bidirected String Graph Model of Genome Assembly

### 2012-11-28 presentation/presentation.tex
presentation fixes

\begin{frame}{Three types of overlaps}
	\begin{enumerate}
		\item ``normal'' overlap
			\begin{center}
				\begin{tikzpicture}
					\draw[->,style=thick] (0,0) -- (2,0);
					\draw[->,style=thick] (1.0,0.3) -- (3.0,0.3);
				\end{tikzpicture}

@@

				\begin{tikzpicture}
					\draw[->,style=thick] (2,0) -- (0,0);
					\draw[->,style=thick] (3.0,0.3) -- (1.0,0.3);
				\end{tikzpicture}
			\end{center}
		\item ``innie'' overlap
			\begin{center} \begin{tikzpicture}
					\draw[->,style=thick] (0,0) -- (2,0);
					\draw[->,style=thick] (3,0.3) -- (1,0.3);
			\end{tikzpicture} \end{center}
		\item ``outtie'' overlap
			\begin{center} \begin{tikzpicture}
					\draw[->,style=thick] (2,0) -- (0,0);
					\draw[->,style=thick] (1.0,0.3) -- (3.0,0.3);
			\end{tikzpicture} \end{center}
	\end{enumerate}
\end{frame}

@@

\begin{frame}{Computing overlaps (Implementation)}
	\begin{itemize}
		\item Implementation: {\tt compute-overlaps}
			\begin{itemize}
				\item {\bf Input}:  A set of reads and the minimum overlap
				length $\ell$
				\item {\bf Output}:  All exact pairwise overlaps between reads,
				including reverse-complement overlaps
				\item {\bf Algorithm}:  Seed-and-extend
				\item {\bf Running time}: Linear with regards to genome length for
				non-repetitive genomes; quadratic with regards to read density;
				quadratic with regards to the number of times a sequence of DNA
				is repeated in the genome
			\end{itemize}
	\end{itemize}
	%\begin{algorithm}[H]
		%\caption{Algorithm {\sc compute-overlaps}($S$)}
		%\begin{algorithmic}
			%%\STATE{h = new hash table}
			%%\FOR{{\bf each} read $r \in S$}
				%%\FOR{{\bf each} canonical $k$-mer $K \in R$}
					%%\STATE{Append $K.pos$ to $h[K]$}
				%%\ENDFOR
			%%\ENDFOR
			%%\FORALL{posList $\in h$}
				%%\FORALL{$k$-mer pairs $k1$, $k2$ in 
			%%\ENDFOR
		%\end{algorithmic}
	%\end{algorithm}
\end{frame}

@@

		\item We must allow each read to be used either in the
		reverse-complement or forward direction.
		\item Therefore, we must allow the edges to be traversed in either
		direction.
		\item However, each read must still be used in a consistent way within
		any single reconstruction.
		\item Solution: use a {\bf bidirected graph}.
	\end{itemize}
\end{frame}

@@

\begin{frame}{Walks on a bidirected graph}
	\begin{itemize}
		\item A {\bf walk on a bidirected graph} is a continuous sequence of
		edges such that if any vertex $v$ is entered through a head inwards, it
		is left on a head outwards (unless it is the end of the path), and vice
		versa.
		\item A bidirected edge may be traversed in opposite directions during
		the same walk.
		\item The reverse of a walk in a directed graph is also a bidirected walk.
		\item $1 \to 2 \to 3 \to 4$ and $4 \to 3 \to 2 \to 1$ are both valid
		walks in the bidirected graph below.
	\end{itemize}
		\begin{center} {\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
			\node (1) at (0.00, 0) {1};
			\node (2) at (1.75, 0) {2};
			\node (3) at (3.50, 0) {3};
			\node (4) at (5.25, 0) {4};
			\draw[<->,very thick] (1) edge (2);
			\draw[>-<,very thick] (2) edge (3);
			\draw[<-<,very thick] (3) edge (4);
		\end{tikzpicture} } \end{center}
\end{frame}

@@

\begin{frame}{Transitive reduction (cont.)}
		Transitive reduction algorithm:  Go through each vertex and examine
		neighbors up to 2 steps away to identify all transitive edges leaving this
		vertex.
	\begin{algorithm}[H]
		\caption{{\sc TransitiveReduction}($G$)}
	{\footnotesize
		\begin{algorithmic}
			\FOR{{\bf each} vertex $v \in G$}
				\STATE{Mark all neighbors of $v$ as {\tt INPLAY}.}
				\STATE{Set {\tt longest} $\gets$ length of the longest edge leaving $v$}
				\FOR{{\bf each} edge $v \to w$ in order of length}
					\FOR{{\bf each} edge $w \to x$ in order of length}
						\IF{$x$ is not marked {\tt INPLAY}}
							\STATE{\bf continue}
						\ENDIF
						\IF{$len(v \to w) + len(w \to x) \ge$ {\tt longest}}
							\STATE{\bf break}
						\ENDIF
						\STATE{Mark $x$ as {\tt ELIMINATED}}
					\ENDFOR
				\ENDFOR
				\STATE{Mark as transitive all edges $v \to x$ where $x$ is marked {\tt ELIMINATED}}
			\ENDFOR
			\STATE{Eliminate all edges marked transitive.}
		\end{algorithmic}
	}
	\end{algorithm}
\end{frame}

@@

	\begin{figure}[H]
		\begin{tikzpicture}
			{\small
			\begin{tikzpicture}[>=triangle 45]
				\node[circle,fill=blue!20,draw=black] (A) at (0.0, 0) {A};
				\node[circle,fill=blue!20,draw=black] (B) at (3.0, 0) {B};
				\draw[>->,very thick] (A) edge node[anchor=north] {4639100} (B);
			\end{tikzpicture} }
		\end{tikzpicture}
		\caption{
			Bidirected string graph for reads uniformly sampled from a 4.6 Mb
			random genome after transitive reduction and path collapsing.  (Edge
			labeled with length of sequence)}
	\end{figure}
\end{frame}

>>>>

\begin{frame}{Three types of overlaps}
	\begin{enumerate}
		\item ``Normal'' overlap
			\begin{center}
				\begin{tikzpicture}
					\draw[->,style=thick] (0,0) -- (2,0);
					\draw[->,style=thick] (1.0,0.3) -- (3.0,0.3);
				\end{tikzpicture}

@@

				\begin{tikzpicture}
					\draw[->,style=thick] (2,0) -- (0,0);
					\draw[->,style=thick] (3.0,0.3) -- (1.0,0.3);
				\end{tikzpicture}
			\end{center}
		\item ``Innie'' overlap
			\begin{center} \begin{tikzpicture}
					\draw[->,style=thick] (0,0) -- (2,0);
					\draw[->,style=thick] (3,0.3) -- (1,0.3);
			\end{tikzpicture} \end{center}
		\item ``Outtie'' overlap
			\begin{center} \begin{tikzpicture}
					\draw[->,style=thick] (2,0) -- (0,0);
					\draw[->,style=thick] (1.0,0.3) -- (3.0,0.3);
			\end{tikzpicture} \end{center}
	\end{enumerate}
\end{frame}

@@

\begin{frame}{Computing overlaps (Implementation)}
	\begin{itemize}
		\item Implementation: {\tt compute-overlaps}
			\begin{itemize}
				\item {\bf Input}:  A set of reads and the minimum overlap
				length $\ell$
				\item {\bf Output}:  All exact pairwise overlaps between reads,
				including reverse-complement overlaps
				\item {\bf Algorithm}:  Seed-and-extend
				\item {\bf Running time}: Linear with respect to genome length
				for non-repetitive genomes; quadratic with respect to read
					density; quadratic with respect to the number of times a
					sequence of DNA is repeated in the genome
			\end{itemize}
	\end{itemize}
	%\begin{algorithm}[H]
		%\caption{Algorithm {\sc compute-overlaps}($S$)}
		%\begin{algorithmic}
			%%\STATE{h = new hash table}
			%%\FOR{{\bf each} read $r \in S$}
				%%\FOR{{\bf each} canonical $k$-mer $K \in R$}
					%%\STATE{Append $K.pos$ to $h[K]$}
				%%\ENDFOR
			%%\ENDFOR
			%%\FORALL{posList $\in h$}
				%%\FORALL{$k$-mer pairs $k1$, $k2$ in 
			%%\ENDFOR
		%\end{algorithmic}
	%\end{algorithm}
\end{frame}

@@

		\item We must allow each read to be used either in the
		reverse-complement or forward direction.
		\item Therefore, we must allow the edges to be traversed in either
		direction.
		\item However, each read must still be used in a consistent way within
		any single assembly.
		\item Solution: use a {\bf bidirected graph}.
	\end{itemize}
\end{frame}

@@

\begin{frame}{Walks in a bidirected graph}
	\begin{itemize}
		\item A {\bf walk in a bidirected graph} $G$ is a continuous sequence of
		edges in $G$ such that if any vertex $v$ is entered through a head
		inwards, it is left on a head outwards (unless it is the end of the
		path), and vice versa.
		\item A bidirected edge may be traversed in opposite directions during
		the same walk.
		\item The reverse of a walk in a directed graph is also a bidirected walk.
		\item $1 \to 2 \to 3 \to 4$ and $4 \to 3 \to 2 \to 1$ are both valid
		walks in the bidirected graph below.
	\end{itemize}
		\begin{center} {\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
			\node (1) at (0.00, 0) {1};
			\node (2) at (1.75, 0) {2};
			\node (3) at (3.50, 0) {3};
			\node (4) at (5.25, 0) {4};
			\draw[<->,very thick] (1) edge (2);
			\draw[>-<,very thick] (2) edge (3);
			\draw[<-<,very thick] (3) edge (4);
		\end{tikzpicture} } \end{center}
\end{frame}

@@

\begin{frame}{Transitive reduction (cont.)}
		Transitive reduction algorithm:  Go through each vertex and examine
		neighbors up to 2 steps away to identify all transitive edges leaving this
		vertex.
	\begin{algorithm}[H]
		\caption{{\sc TransitiveReduction}($G$)}
	{\footnotesize
		\begin{algorithmic}
			\FOR{{\bf each} vertex $v \in G$}
				\STATE{Mark all neighbors of $v$ as {\tt INPLAY}.}
				\STATE{Set {\tt longest} $\gets$ length of the longest edge leaving $v$}
				\FOR{{\bf each} edge $v \to w$ in order of length}
					\FOR{{\bf each} edge $w \to x$ in order of length}
						\IF{$x$ is not marked {\tt INPLAY}}
							\STATE{\bf continue}
						\ENDIF
						\IF{$len(v \to w) + len(w \to x) > $ {\tt longest}}
							\STATE{\bf break}
						\ENDIF
						\STATE{Mark $x$ as {\tt ELIMINATED}}
					\ENDFOR
				\ENDFOR
				\STATE{Mark as transitive all edges $v \to x$ where $x$ is marked {\tt ELIMINATED}}
			\ENDFOR
			\STATE{Eliminate all edges marked transitive.}
		\end{algorithmic}
	}
	\end{algorithm}
\end{frame}

@@

	\begin{figure}[H]
		\begin{tikzpicture}
			{\small
			\begin{tikzpicture}[>=triangle 45]
				\node[circle,fill=blue!20,draw=black] (A) at (0.0, 0) {A};
				\node[circle,fill=blue!20,draw=black] (B) at (3.0, 0) {B};
				\draw[>->,very thick] (A) edge node[anchor=north] {4639100} (B);
			\end{tikzpicture} }
		\end{tikzpicture}
		\caption{
			Bidirected string graph for reads uniformly sampled from a 4.6 Mbp
			random genome after transitive reduction and path collapsing.  (Edge
			labeled with length of sequence)}
	\end{figure}
\end{frame}
